---
title: "Predictive Analysis of Employee Attrition using Machine Learning"
author: "Data Science Student"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: readable
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Objective

The primary objective of this project is to analyze human resources (HR) data to uncover the driving factors behind employee turnover and to build a predictive machine learning model capable of identifying employees who are at high risk of leaving the organization.

## 2. Problem Statement

High employee turnover entails substantial organizational costs, including expenses related to recruitment, onboarding, training, and the temporary loss of productivity. Understanding *why* employees leave and successfully predicting *who* will leave allows HR departments to deploy targeted proactive retention strategies. 

This project aims to answer two key questions:
1. Which variables (e.g., salary, overtime, job satisfaction) most heavily influence attrition?
2. Can we reliably predict employee attrition using a statistical machine learning model?

## 3. Dataset Description

Because real HR data is often highly confidential, a realistic synthetic dataset consisting of 1,000 employee records is generated for this project. The dataset encompasses various demographic and job-related variables:

* **Age** (Numeric): The employee's age in years (Range: 18 - 65).
* **Department** (Categorical): The corporate department (Sales, R&D, HR).
* **JobSatisfaction** (Ordinal): Self-reported satisfaction score on a scale from 1 (Low) to 4 (High).
* **OverTime** (Categorical): Indicates whether the employee regularly works overtime (Yes/No).
* **MonthlyIncome** (Numeric): The employee's estimated monthly salary in USD.
* **YearsAtCompany** (Numeric): The number of years the employee has spent at the company.
* **Attrition** (Target Variable): Indicates whether the employee has left the company (Yes/No).

---

## 4. Methodology and Data Generation

### Loading Required Libraries

```{r load-libraries}
packages <- c("dplyr", "tidyr", "ggplot2", "caret", "e1071")
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages], repos = "https://cran.r-project.org/")
}
library(dplyr)
library(ggplot2)
library(caret)
```

### Generating the Dataset

Below, we generate the synthetic HR dataset.

```{r generate-data}
set.seed(42) # For reproducibility
n <- 1000

# Generating features
Age <- pmax(18, pmin(round(rnorm(n, mean = 38, sd = 10)), 65))
Department <- sample(c("Sales", "Research & Development", "Human Resources"), n, replace = TRUE, prob = c(0.3, 0.6, 0.1))
JobSatisfaction <- sample(1:4, n, replace = TRUE, prob = c(0.15, 0.2, 0.35, 0.3))
OverTime <- sample(c("Yes", "No"), n, replace = TRUE, prob = c(0.28, 0.72))

# Income depends on department
MonthlyIncome <- ifelse(Department == "Sales", rnorm(n, 5500, 1500),
                 ifelse(Department == "Research & Development", rnorm(n, 6500, 2000),
                        rnorm(n, 4800, 1200)))
MonthlyIncome <- round(abs(MonthlyIncome)) + 2000
YearsAtCompany <- round(runif(n, min = 0, max = Age - 18))

# Simulating Attrition based on a logistic function
logit_score <- 1.5 - 0.04 * Age + 1.8 * (OverTime == "Yes") - 0.6 * JobSatisfaction - 0.00015 * MonthlyIncome + 0.05 * YearsAtCompany
logit_score <- logit_score + rnorm(n, 0, 1) # random noise
prob_attrition <- 1 / (1 + exp(-logit_score))
Attrition <- ifelse(prob_attrition > 0.50, "Yes", "No")

# Combine into a dataframe
hr_data <- data.frame(Age, Department, JobSatisfaction, OverTime, MonthlyIncome, YearsAtCompany, Attrition)
```

### Data Preprocessing

We verify the absence of missing values and properly format categorical features.

```{r data-prep}
# Check for missing values
sum(is.na(hr_data))

# Format data types
hr_data <- hr_data %>%
  mutate(
    Department = as.factor(Department),
    JobSatisfaction = as.factor(JobSatisfaction), # Factor for categorical plotting/modeling
    OverTime = as.factor(OverTime),
    Attrition = as.factor(Attrition)
  )

summary(hr_data)
```

---

## 5. Exploratory Data Analysis (EDA)

Visualizing the data helps us understand the relationships between different attributes and our target variable `Attrition`.

### Target Distribution

```{r eda-attrition, fig.width=6, fig.height=4}
ggplot(hr_data, aes(x = Attrition, fill = Attrition)) +
  geom_bar() +
  scale_fill_manual(values = c("No" = "#4CAF50", "Yes" = "#F44336")) +
  labs(title = "Distribution of Employee Attrition", x = "Attrition", y = "Employee Count") +
  theme_minimal()
```
*Insight*: The dataset is slightly imbalanced, which is highly typical for human resources data (most employees stay).

### The Impact of Overtime

```{r eda-overtime, fig.width=6, fig.height=4}
ggplot(hr_data, aes(x = OverTime, fill = Attrition)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Attrition by OverTime", x = "OverTime", y = "Percentage") +
  theme_minimal()
```
*Insight*: There is a substantially higher percentage of attrition among employees who work overtime compared to those who do not.

### Age and Monthly Income Dynamics

```{r eda-age-income, fig.width=10, fig.height=4}
# Income Density
plot_inc <- ggplot(hr_data, aes(x = MonthlyIncome, fill = Attrition)) +
  geom_density(alpha = 0.5) +
  labs(title = "Monthly Income Density by Attrition", x = "Monthly Income", y = "Density") +
  theme_minimal()

# Age Boxplot
plot_age <- ggplot(hr_data, aes(x = Attrition, y = Age, fill = Attrition)) +
  geom_boxplot() +
  labs(title = "Employee Age vs Attrition", x = "Attrition", y = "Age") +
  theme_light()

gridExtra::grid.arrange(plot_inc, plot_age, ncol=2)
```
*Insight*: Lower monthly incomes and younger age brackets visibly skew towards a higher likelihood of attrition.

---

## 6. Implementation of Machine Learning Model

We will construct a **Logistic Regression** model since our target is a binary categorical outcome.

### Train-Test Split

```{r model-split}
set.seed(123)
# 70% Training / 30% Testing Split
trainIndex <- createDataPartition(hr_data$Attrition, p = 0.7, list = FALSE, times = 1)
data_train <- hr_data[trainIndex, ]
data_test  <- hr_data[-trainIndex, ]
```

### Model Training

```{r model-train}
logit_model <- glm(Attrition ~ Age + Department + JobSatisfaction + OverTime + MonthlyIncome + YearsAtCompany, 
                   data = data_train, 
                   family = binomial)

summary(logit_model)
```
*Analysis of Model Output*: The p-values `Pr(>|z|)` indicate that `Age`, `OverTime`, `JobSatisfaction`, and `MonthlyIncome` are statistically significant predictors at the `0.05` alpha level.

### Model Evaluation

```{r model-eval}
# Generate probability predictions
pred_prob <- predict(logit_model, newdata = data_test, type = "response")

# Convert to Class Labels (Decision Threshold: 50%)
pred_class <- as.factor(ifelse(pred_prob > 0.5, "Yes", "No"))
levels(pred_class) <- levels(data_test$Attrition)

# Confusion Matrix
conf_matrix <- confusionMatrix(pred_class, data_test$Attrition)
conf_matrix
```

---

## 7. Results and Conclusion

### Key Findings
1. **Model Efficacy**: The logistic regression model establishes an accuracy of approximately **`r round(conf_matrix$overall['Accuracy'] * 100, 2)`%**, which is a strong baseline.
2. **Actionable Variables**: `OverTime`, `JobSatisfaction`, and Salary bands represent actionable items for HR departments. Overworked and underpaid associates exhibit the highest risk.

### Business Recommendations
The logistic regression model provides HR with a pragmatic tool to flag employees exhibiting "high-risk" profiles. Reallocating budget to retention programs (like targeted raises, restricting overtime, or addressing workplace morale) for flagged employees can result in substantial cost savings.

## 8. Suggestions for Future Improvement
*   **Feature Engineering**: Incorporating features like literal commute time, performance rating, and survey scores from peers.
*   **Tree-based Ensembles**: Trying non-linear classifiers like Random Forest or XGBoost.
*   **Cross Validation**: Using k-fold cross-validation to guarantee the robustness of the accuracy metrics.

## 9. References
*   R Core Team (2023). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria.
*   Wickham, H., et al. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, 4(43), 1686.
*   Kuhn, M. (2008). Building Predictive Models in R Using the caret Package. *Journal of Statistical Software*, 28(5), 1-26.
